Abstractions and exemplars: the measure noun phrase alternation in German

### 2018-04-21 Latex/sections/1intro.tex
First half of proofreading.

This paper deals with a morpho-syntactic alternation between two measure noun constructions in German.
By \textit{alternation} I refer to a situation where two or more forms or constructions are available with no clear difference in acceptability, function, or meaning.
The study of lexical and constructional alternations has a long history in cognitive corpus linguistics (for example, \citealp{BresnanEa2007,BresnanHay2010,BresnanFord2010,DivjakArppe2013,Gries2015,NessetJanda2010}).
This research is based on the assumption that language is a probabilistic phenomenon where alternants are chosen neither deterministically nor fully at random \citep{Bresnan2007}.
Multifactorial models are constructed which incorporate influencing factors from diverse levels, including lexical and contextual factors.
The estimation of the model coefficients quantifies the influence that the factors have on the probability that either alternant is chosen.
There are two fundamental issues to consider with respect to this tradition.
First, there is the question of whether corpus data provide any insight into cognitive representations at all.
This question can and should be answered by testing how well corpus-derived models converge with or diverge from experimental findings (Secion~\ref{sec:usagedataincognitivelinguistics}).
Second, the appropriate modelling of such results in cognitive linguistics -- \ie\ the assumed constructs -- is a key issue (Section~\ref{sec:prototypesexemplarscorporaandcontrolledexperiments}).

@@

For some time, there has been an interest in correlating probabilistic generalisations extracted from corpus data and results from experimental work (for example, \citealp{ArppeJaervikivi2007,BresnanEa2007,BresnanFord2010,DivjakGries2008,DivjakEa2016,FordBresnan2013}).
This is often called a \textit{validation} of the corpus-derived findings, but \citet[303]{Divjak2016a} criticises this choice of words ``because it creates the impression that behavioral experimental data is inherently more valuable than textual data'', citing \cite{TummersEa2005}, who state that a corpus is ``a sample of spontaneous language use that is (generally) realized by native speakers''.
See also \citet{Newman2011} for a positive view of corpora as a source of data in cognitive linguistics in their own right.
However, as \citet[486--487]{Dabrowska2016} argues, this does not mean that we can in some way ``deduce mental representations from patterns of use'', \ie\ from corpus data.
It would indeed be highly surprising if this were possible, and the same holds for experimental methods.
Nobody assumes that we can inductively infer mental representations from experiments, which even allow for direct access to the cognitive agent and offer much better ways of controlling experimental conditions and nuisance variables.
Rather, predictions are derived from existing theories in order to \textit{test} the theories.
While this approach is applicable to corpus data, I discuss some relevant differences.

As mentioned above, the central question is whether usage data as found in corpora are truly predictive of speakers' and writers' cognitive representation of language and\slash or of their linguistic behaviour.
This is where the experimental validation (or \textit{corroboration}) comes into play.
A review of the state of the art was provided by \cite{NewmanSorensenduncan2015}, who enumerate a number of studies showing how corpus data and experimental data converge (such as \citealp{BresnanEa2007,DurrantDoherty2010,GriesWulff2005,GriesEa2005}) and a number of studies where the two types of data led to diverging or only partially converging results (such as \citealp{ArppeJaervikivi2007,Dabrowska2014,Mollin2009}).
When researchers do not achieve convergence, they often try to explain this by differentiating between the actual cognitive construct and whatever the pooled usage data as found in corpora represent.
For example, \citet[411]{Dabrowska2014} lists a number of possible reasons to explain why subjects in her experiment diverged in their word association preferences from collocation measures extracted from corpora.
Alternatively, researchers argue for a more adequate statistical analysis to increase the fit between corpus data and experimental data.
See, for example \cite{DivjakEa2016}, who show that generalised additive models (GAMs) are better suited than generalised linear models (GLMs) for correlating reading times and corpus data. 
Much more optimistically, \cite{StefanowitschFlach2016} proposed a straightforward positivist perspective of corpora as representing the input of an average adult speaker, thus licensing inferences from corpus data to cognitive representations under a ``corpus-as-input'' view.
They state that ``in this wider context, large, register-mixed corpora such as the British National Corpus [\ldots] may not be perfect models of the linguistic experience of adult speakers, but they are reasonably close to the input of an idealized average member of the relevant speech community'' \citep[104]{StefanowitschFlach2016}.
In essence, this entails that any generalisation extracted from the BNC can be assumed to have some kind of mental reality, which is at least doubtful.

@@

In cognitive linguistics, similarity-based categorisation is often seen as the central and conceptually sufficient approach to linguistic categorisation.
See \cite{Taylor2003} for a comprehensive treatment in terms of prototype theory and \cite{Taylor2012} for a detailed discussion of an exemplar-based approach.
\cite{DivjakArppe2013} is a rare example of a paper in cognitive linguistics where a synthesis between prototype and exemplar models is proposed.
Their corpus-based approach shows ``one way of systematically analyzing usage data as contained in corpora to yield a scheme, compatible with usage-based theories of language, by which the assumptions of both the prototype and exemplar theories can be operationalized'' \citep[267]{DivjakArppe2013}.
Their approach to implementing a varying abstraction model \citep[254--260]{DivjakArppe2013} is based on hierarchical clustering of annotated properties of sentences.
They cluster sentences containing Russian verbs of trying.
Then, they single out the one sentence from each cluster which scores the highest probability for any of the six \textit{try} verbs according to a polytomous regression model estimated on the same data.
The clusters are interpreted as intermediate-level exemplar-derived abstractions of typical contexts for these high-probability verbs (typically more than one cluster for each verb; \citealp[255--256]{DivjakArppe2013}).
The crucial difference between such data-driven corpus-based analyses and experiments in cognitive science (\citealp{DivjakArppe2013} use \citealp{VerbeemenEa2007} as their reference) is that cognitive research is based on experiments where subjects produce actual category assignments or similarity judgements, and in corpus studies, the categories and category membership are determined purely from existing data.
The experimental approach with reduced and\slash or artificial stimuli makes it much easier to examine very specific effects in the behaviour of the subjects.
While I am convinced that the results presented in \citet{DivjakArppe2013} are valid and important (especially given the previous and subsequent research the authors have conducted on the data, including experimental work presented in \citealp{DivjakEa2016}), any data set can be clustered to yield a certain number of clusters.
Thus, the study does not ensure that the clusters emerging from the data correspond to any speaker's cognitive representation.

On the other hand, the trade-off one has to accept when doing experiments with highly simplified stimuli is lower \textit{external validity} (\ie\ decreased generalisability) and higher dependence on proper operationalisations of constructs (\textit{construct validity}).%
\footnote{An accessible overview of the different types of validity can be found in Chapter~1 of \citet{MaxwellDelaney2004}.}
Tasks in cognitive science have been criticised exactly for their lack of external validity, for example by \citet{Murphy2003}.
From a linguistic perspective, it seems remarkable that \citet[1013]{VoorspoelsEa2011} consider their experimental task -- the assignment of typicality scores to nouns from the domains of \textit{animals} and \textit{artefacts} to categories like \textit{bird}, \textit{fish}, \textit{clothing}, or \textit{tools} -- a study of ``superordinate natural language categories, whereas most evidence supporting exemplar representations has been found in artificial categories of a more subordinate level''.
Corpus linguists interested in probabilistic alternation modelling investigate significantly more complex high-level categories and use large and complex feature sets.%
\footnote{However, approaches have emerged which model linguistic phenomena without reference to high-level linguistic features altogether \citep{BaayenEa2016,RamscarPort2016}.}
It is thus an advantage of much linguistic work on categorisation that it deals with complex and realistically produced data because this greatly improves the external validity of studies, although by potentially sacrificing some construct validity.
Thus, an ideal contribution to the research on category abstraction by cognitive corpus linguists is to provide analyses which have great external validity and complexity while carefully making sure that these finding correlate with actions and reactions under more controlled experimental conditions, thus increasing the construct validity.

This paper contributes to solving the questions raised in this section in many ways.
After a thorough description of the alternation under examination, I discuss potential influencing factors comprising high-level abstract semantic generalisations as well as exemplar-similarity and item-specific effects in Section~\ref{sec:germanmeasurenps}.
I will argue that there are lemma-specific and exemplar effects but also generalisations at the level of the construction as well as generalisations overlaying the lemma-specific effects, leading to a complex hierarchical structure.
In Section~\ref{sec:corpusstudies}, I present a corpus study and report a true multilevel generalised linear model with the appropriate hierarchical structure for the hypothesised effects.
In Section~\ref{sec:experimental}, I test the predictions of the corpus-based model in two experimental paradigms (forced-choice and self-paced reading), showing that they indeed converge, albeit with low effect strength.
In Section~\ref{sec:conclusion}, I interpret the findings in the light of the issues of cognitive representations and corpus data and the convergence of usage data and experimental data.

>>>>

This paper deals with a morpho-syntactic alternation between two measure noun constructions in German.
By \textit{alternation} I refer to a situation where two or more forms or constructions are available with no clear difference in acceptability, function, or meaning.
The study of lexical and constructional alternations has a long history in cognitive corpus linguistics (for example, \citealp{BresnanEa2007,BresnanHay2010,BresnanFord2010,DivjakArppe2013,Gries2015,NessetJanda2010}).
This research is based on the assumption that language is a probabilistic phenomenon where alternants are chosen neither deterministically nor fully at random \citep{Bresnan2007}.
Multifactorial models are constructed which incorporate influencing factors from diverse levels, including lexical and contextual factors.
The estimation of the model coefficients quantifies the influence that the factors have on the probability that either alternant is chosen.
There are two fundamental issues to consider with respect to this tradition.
First, there is the question of whether corpus data provide any insight into cognitive representations at all.
This question can and should be answered by testing how well corpus-derived models converge with or diverge from experimental findings (Secion~\ref{sec:usagedataincognitivelinguistics}).
Second, the appropriate modelling of such results in cognitive linguistics is a key issue (Section~\ref{sec:prototypesexemplarscorporaandcontrolledexperiments}).

@@

For some time, there has been an interest in correlating probabilistic generalisations extracted from corpus data and results from experimental work \citep{ArppeJaervikivi2007,BresnanEa2007,BresnanFord2010,DivjakGries2008,DivjakEa2016,FordBresnan2013}.
This is often called a \textit{validation} of the corpus-derived findings, but \citet[303]{Divjak2016a} criticises this choice of words ``because it creates the impression that behavioral experimental data is inherently more valuable than textual data'', citing \cite{TummersEa2005}, who state that a corpus is ``a sample of spontaneous language use that is (generally) realized by native speakers''.
See also \citet{Newman2011} for a positive view of corpora as a source of data in cognitive linguistics in their own right.
However, as \citet[486--487]{Dabrowska2016} argues, this does not mean that we can in some way ``deduce mental representations from patterns of use'', \ie\ from corpus data.
It would indeed be highly surprising if this were possible, and the same holds for experimental methods.
Nobody assumes that we can inductively infer mental representations from experiments, which even allow for direct access to the cognitive agent and offer much better ways of controlling experimental conditions and nuisance variables.
Rather, predictions are derived from existing theories in order to \textit{test} the theories.
While this approach is applicable to corpus data, I discuss some relevant differences below.

As mentioned above, the central question is whether usage data as found in corpora are truly predictive of speakers' and writers' cognitive representation of language and\slash or of their linguistic behaviour.
This is where the experimental validation (or \textit{corroboration}) comes into play.
A review of the state of the art was provided by \cite{NewmanSorensenduncan2015}, who enumerate a number of studies showing how corpus data and experimental data converge (such as \citealp{BresnanEa2007,DurrantDoherty2010,GriesWulff2005,GriesEa2005}) and a number of studies where the two types of data led to diverging or only partially converging results (such as \citealp{ArppeJaervikivi2007,Dabrowska2014,Mollin2009}).
When researchers do not achieve convergence, they often try to explain this by differentiating between the actual cognitive construct and whatever the pooled usage data as found in corpora represent.
For example, \citet[411]{Dabrowska2014} lists a number of possible reasons to explain why subjects in her experiment diverged in their word association preferences from collocation measures extracted from corpora.
Alternatively, researchers argue for a more adequate statistical analysis to increase the fit between corpus data and experimental data.
See, for example \cite{DivjakEa2016}, who show that generalised additive models (GAMs) are better suited than generalised linear models (GLMs) for correlating reading times and corpus data. 
Much more optimistically, \cite{StefanowitschFlach2016} proposed a straightforward positivist perspective of corpora as representing the input of an average adult speaker, thus licensing inferences from corpus data to cognitive representations under a ``corpus-as-input'' view.
They state that ``in this wider context, large, register-mixed corpora such as the British National Corpus [\ldots] may not be perfect models of the linguistic experience of adult speakers, but they are reasonably close to the input of an idealized average member of the relevant speech community'' \citep[104]{StefanowitschFlach2016}.
If this were a valid assumption then any generalisation extracted from the BNC could be assumed to have some kind of mental reality, which is doubtful.

@@

In cognitive linguistics, similarity-based categorisation is often seen as the central and conceptually sufficient approach to linguistic categorisation.
See \cite{Taylor2003} for a comprehensive treatment in terms of prototype theory and \cite{Taylor2012} for a detailed discussion of an exemplar-based approach.
\cite{DivjakArppe2013} is a rare example of a paper in cognitive linguistics where a synthesis between prototype and exemplar models is proposed.
Their corpus-based approach shows ``one way of systematically analyzing usage data as contained in corpora to yield a scheme, compatible with usage-based theories of language, by which the assumptions of both the prototype and exemplar theories can be operationalized'' \citep[267]{DivjakArppe2013}.
Their approach to implementing a varying abstraction model \citep[254--260]{DivjakArppe2013} is based on hierarchical clustering of annotated properties of sentences.
They cluster sentences containing Russian verbs of trying.
Then, they single out the one sentence from each cluster which scores the highest probability for any of the six \textit{try} verbs according to a polytomous regression model estimated on the same data.
The clusters are interpreted as intermediate-level exemplar-derived abstractions of typical contexts for these high-probability verbs (typically more than one cluster for each verb; \citealp[255--256]{DivjakArppe2013}).
The crucial difference between such data-driven corpus-based analyses and experiments in cognitive science (\citealp{DivjakArppe2013} use \citealp{VerbeemenEa2007} as their reference) is that cognitive research is based on experiments where subjects produce actual category assignments or similarity judgements, and in corpus studies, the categories and category membership are determined purely from existing data.
The experimental approach with reduced and\slash or artificial stimuli makes it much easier to examine very specific effects in the behaviour of the subjects.
While I am convinced that the results presented in \citet{DivjakArppe2013} are valid and important (especially given the previous and subsequent research the authors have conducted on the data, including experimental work presented in \citealp{DivjakEa2016}), any data set can be clustered to yield a certain number of clusters.
Thus, the study does not ensure that the clusters emerging from the data correspond to any particular speaker's cognitive representation.

On the other hand, the trade-off one has to accept when doing experiments with highly simplified stimuli is lower \textit{external validity} (\ie\ decreased generalisability) and higher dependence on proper operationalisations of constructs (\textit{construct validity}).%
\footnote{An accessible overview of the different types of validity can be found in Chapter~1 of \citet{MaxwellDelaney2004}.}
Tasks in cognitive science have been criticised exactly for their lack of external validity, for example by \citet{Murphy2003}.
From a linguistic perspective, it seems remarkable that \citet[1013]{VoorspoelsEa2011} consider their experimental task -- the assignment of typicality scores to nouns from the domains of \textit{animals} and \textit{artefacts} to categories such as \textit{bird}, \textit{fish}, \textit{clothing}, or \textit{tools} -- a study of ``superordinate natural language categories, whereas most evidence supporting exemplar representations has been found in artificial categories of a more subordinate level''.
Corpus linguists interested in probabilistic alternation modelling investigate significantly more complex high-level categories and use large and complex feature sets.%
\footnote{However, approaches have emerged which model linguistic phenomena without reference to high-level linguistic features altogether \citep{BaayenEa2016,RamscarPort2016}.}
It is thus an advantage of much linguistic work on categorisation that it deals with complex and realistically produced data because this greatly improves the external validity of studies, although by potentially sacrificing some construct validity.
Thus, the ideal contribution to the research on category abstraction by cognitive corpus linguists is to provide analyses which have high external validity and complexity while carefully making sure that these finding correlate with actions and reactions under more controlled experimental conditions, thus increasing the construct validity.

This paper contributes to solving the questions raised in this section in many ways.
After a thorough description of the alternation under examination, I discuss potential influencing factors comprising high-level abstract semantic generalisations as well as exemplar-similarity and item-specific effects in Section~\ref{sec:germanmeasurenps}.
I will argue that there are lemma-specific and exemplar effects but also generalisations at the level of the construction as well as generalisations overlaying the lemma-specific effects, leading to a complex hierarchical structure.
In Section~\ref{sec:corpusstudies}, I present a corpus study and report a true multilevel generalised linear model with the appropriate hierarchical structure for the hypothesised effects.
In Section~\ref{sec:experimental}, I test the predictions of the corpus-based model in two experimental paradigms (forced-choice and self-paced reading), demonstrating that they indeed converge, albeit with low effect strength.
In Section~\ref{sec:conclusion}, I interpret the findings in the light of the issues of cognitive representations and corpus data and the convergence of usage data and experimental data.

### 2018-04-21 Latex/sections/2descriptive.tex
First half of proofreading.

In this section, I introduce and illustrate the relevant alternating constructions in Section~\ref{sec:descriptive}.
Then, I develop a model with prototype effects as well as exemplar effects based in existing research and my own theorising in Section~\ref{sec:analyses}.

@@

In order to understand the MNP alternation, it is vital to consider how in the relevant syntactic structure [N\Sub{1}~[A~N\Sub{2}]\Sub{NP\Sub{2}}]\Sub{NP\Sub{1}} (as instantiated in the \NACa\ and the \PGCa), the adjective is morpho-syntactically ambiguous between a determiner and a modifying adjective.
To show this, the so-called strong and weak inflection patterns of German adjectives needs to be taken into account.
In NPs with a strongly inflected determiner, attributive adjectives inflect according to the massively syncretic \textit{weak} pattern.
If there is no determiner (as is the case in the alternating constructions), attributive adjectives inflect like determiners themselves.
This is called the \textit{strong} inflectional pattern.
For example, in the dative (governed here by the preposition \textit{mit}), the strong suffix \textit{-em} is on the determiner in (\ref{ex:starkschwach:a}) and the adjective bears the weak suffix \textit{-en}. 
In (\ref{ex:starkschwach:b}), the strong suffix \textit{-em} appears on the adjective because there is no determiner.%
\footnote{In the masculine and neuter singular genitive, the strong and weak forms are indistinguishable.
Since the alternation itself does not occur in the genitive (see Section~\ref{sec:descriptive}), this does not create any complications for the present study.}

@@

Thus, the adjectives in the \NACa\ and the \PGCa\ have properties of adjectives as well as determiners.
On the one hand, they are lexical adjectives and function as attributive modifiers.
On the other hand, they are inflected like determiners, and they are the leftmost element in the NP, which is typical of determiners.
This unusual double nature of adjectives in NPs without determiners leads to a plausible interpretation of the pattern shown in Table~\ref{tab:constructions}.
If the adjective is classified as a determiner by virtue of carrying the inflectional markers which are otherwise characteristic of determiners, the \PGCa\ is the appropriate choice.
However, if it is classified as an adjective, the \NACa\ suggests itself because of the lack of a determiner.%
\footnote{The generative analysis presented in \cite{Bhatt1990} comes close to this interpretation by analysing the kind NP in the PGC as a DP and in the NAC as an NP.}
This morpho-syntactic ambiguity means that the \NACa\ is in fact a \NACb\ in disguise, and the \PGCa\ is a \PGCd\ in disguise.
This explains why the alternation arises in the first place.
In Section~\ref{sec:analyses}, I will argue that the NAC and the PGC constructions also have semantically distinguishable prototypes, and that the alternation between \PGCa\ and \NACa\ is an alternation between these prototypes.

@@

\begin{exe}
  \ex\label{ex:prototypes} 
  \begin{xlist}
    \ex{\label{ex:prototypes1}\gll Sie nahmen [einen Löffel [irgendeiner Medizin]\Sub{Gen}]\Sub{Acc}.\\
    they took a spoon some medication \\
    \trans They took a spoon(ful) of some medication.
  }
    \ex{\label{ex:prototypes2}\gll Sie kauften [drei Liter [Öl]\Sub{Acc\slash caseless}]\Sub{Acc}.\\
    they bought three liters oil \\
    \trans They bought three liters of oil.
  }
  \end{xlist}
\end{exe}

@@

I begin by showing how the two meanings of the constructions emerge as a consequence of grammaticalisation processes of partitives and similar constructions.
It is often assumed that pseudo-partitives and quantity constructions arise as a form of grammaticalised partitives (\eg\ \citealp[536--539]{Koptjevskaja2001} for Finnish and Estonian, \citealp[559]{Koptjevskaja2001} for European languages in general).
The grammaticalisation path described by \citet[esp.\ 526--530]{Koptjevskaja2001} can start out (in some languages) with constructions involving two referential nouns (not even necessarily forming a single and contiguous NP) and a \textit{separative} meaning as in \textit{(cut) two slices from the cake} \citep[535]{Koptjevskaja2001}.
In this type of construction, it is most obvious that two separate referents (in the given example the cake and the slices) are conceptualised.
The \textit{part-of} meaning of true partitives as in \textit{a slice of the cake} represents the first stage of a development wherein the measure noun can already lose some semantic content, especially when words like \textit{bite} are no longer necessarily interpreted as a piece literally bitten out of something.
The pseudo-partitive stage finally instantiates a \textit{quantity-of} relation, potentially even leading to fully grammaticalised quantifiers such as \textit{a lot}.
In German, the two (now clearly distinct) available constructions have emerged diachronically from a single source through a complex reanalysis process, and the PGC is clearly the older construction \citep[2--4]{Zimmer2015}.
As predicted by the grammaticalisation pattern just described, it still has the potential to form a true partitive (if the kind noun is definite).
Conversely, the NAC lacks this ability to form true partitives and has gone further down the grammaticalisation path.
It is thus not surprising that it has lost (at least prototypically) the semantics which allows both the measure and the substance to be conceptually accessible as independent referents.
As a consequence, we should expect a tendency for the NAC constructions to host more strongly grammaticalised measure nouns.
For example, highly grammaticalised non-referential nouns like \textit{Gramm} `gram' and \textit{Meter} `metre' should occur proportionally more often in NAC constructions than in PGC constructions.
If such preferences can actually be shown in usage patterns, it would lend strong support for the hypothesised difference in the meanings of the prototypes.
In the corpus study, measure lemmas will therefore be annotated with appropriate semantic class labels to check whether semantic classes of measure nouns have different affinities to the two variants.
The actual classification used is based on the list in \citet[530]{Koptjevskaja2001}, but due to the low token frequencies in many of the potential classes, a very coarse classification was used in the end.
With typical examples, the classes are:
\textit{Physical} (typically non-referential precisely measurable units such as \textit{Liter} `litre', \textit{Meter} `metre', \textit{Gramm} `gram'),
\textit{Container} (\textit{Tasse} `cup'),
\textit{Amount} (\textit{Menge} `amount'), 
\textit{Portion} (natural portions like \textit{Happen} `bite' or \textit{Krümel} `crumb').
Notice that \textit{Container} is the class containing words like \textit{spoon} or \textit{cup}, which often develop into partially grammaticalised physical measure nouns.
The lemmas that did not fit into either of these classes were labelled \textit{Rest}.

>>>>

In this section, I introduce and illustrate the relevant alternating constructions in Section~\ref{sec:descriptive}.
Then, I develop a model with prototype effects as well as exemplar effects based on existing research and my own theory in Section~\ref{sec:analyses}.

@@

In order to understand the MNP alternation, it is vital to consider how in the relevant syntactic structure [N\Sub{1}~[A~N\Sub{2}]\Sub{NP\Sub{2}}]\Sub{NP\Sub{1}} (as instantiated in the \NACa\ and the \PGCa), the adjective is morpho-syntactically ambiguous between a determiner and a modifying adjective.
To show this, the so-called strong and weak inflection patterns of German adjectives needs to be taken into account.
In NPs with a strongly inflected determiner, attributive adjectives inflect according to the massively syncretic \textit{weak} pattern.
If there is no determiner (as is the case in the alternating constructions), attributive adjectives inflect as determiners themselves.
This is called the \textit{strong} inflectional pattern.
For example, in the dative (governed here by the preposition \textit{mit}), the strong suffix \textit{-em} is on the determiner in (\ref{ex:starkschwach:a}) and the adjective bears the weak suffix \textit{-en}. 
In (\ref{ex:starkschwach:b}), the strong suffix \textit{-em} appears on the adjective because there is no determiner.%
\footnote{In the masculine and neuter singular genitive, the strong and weak forms are indistinguishable.
Since the alternation itself does not occur in the genitive (see Section~\ref{sec:descriptive}), this does not create any complications for the present study.}

@@

Thus, the adjectives in the \NACa\ and the \PGCa\ have properties of adjectives as well as determiners.
On the one hand, they are lexical adjectives and function as attributive modifiers.
On the other hand, they are inflected as are determiners, and they are the leftmost element in the NP, which is typical of determiners.
This unusual double nature of adjectives in NPs without determiners leads to a plausible interpretation of the pattern shown in Table~\ref{tab:constructions}.
If the adjective is classified as a determiner by virtue of carrying the inflectional markers which are otherwise characteristic of determiners, the \PGCa\ is the appropriate choice.
However, if it is classified as an adjective, the \NACa\ suggests itself because of the lack of a determiner.%
\footnote{The generative analysis presented in \cite{Bhatt1990} comes close to this interpretation by analysing the kind NP in the PGC as a DP and in the NAC as an NP.}
This morpho-syntactic ambiguity means that the \NACa\ is in fact a \NACb\ in disguise, and the \PGCa\ is a \PGCd\ in disguise.
This explains why the alternation arises in the first place.
In Section~\ref{sec:analyses}, I will argue that the NAC and the PGC constructions also have semantically distinguishable prototypes, and that the alternation between \PGCa\ and \NACa\ is an alternation between these prototypes.

@@

\begin{exe}
  \ex\label{ex:prototypes} 
  \begin{xlist}
    \ex{\label{ex:prototypes1}\gll Sie nahmen [einen Löffel [irgendeiner Medizin]\Sub{Gen}]\Sub{Acc}.\\
    they took a spoon some medication \\
    \trans They took a spoon(ful) of any old medication.
  }
    \ex{\label{ex:prototypes2}\gll Sie kauften [drei Liter [Öl]\Sub{Acc\slash caseless}]\Sub{Acc}.\\
    they bought three liters oil \\
    \trans They bought three liters of oil.
  }
  \end{xlist}
\end{exe}

@@

I begin by showing how the two meanings of the constructions emerge as a consequence of grammaticalisation processes of partitives and similar constructions.
It is often assumed that pseudo-partitives and quantity constructions arise as a form of grammaticalised partitives (\eg\ \citealp[536--539]{Koptjevskaja2001} for Finnish and Estonian, \citealp[559]{Koptjevskaja2001} for European languages in general).
The grammaticalisation path described by \citet[esp.\ 526--530]{Koptjevskaja2001} can start out (in some languages) with constructions involving two referential nouns (not even necessarily forming a single and contiguous NP) and a \textit{separative} meaning as in \textit{(cut) two slices from the cake} \citep[535]{Koptjevskaja2001}.
In this type of construction, it is most obvious that two separate referents (in the given example the cake and the slices) are conceptualised.
The \textit{part-of} meaning of true partitives as in \textit{a slice of the cake} represents the first stage of a development wherein the measure noun can already lose some semantic content, especially when words like \textit{bite} are no longer necessarily interpreted as a piece literally bitten out of something.
The pseudo-partitive stage finally instantiates a \textit{quantity-of} relation, potentially even leading to fully grammaticalised quantifiers such as \textit{a lot}.
In German, the two (now clearly distinct) available constructions have emerged diachronically from a single source through a complex reanalysis process, and the PGC is clearly the older construction \citep[2--4]{Zimmer2015}.
As predicted by the grammaticalisation pattern just described, it still has the potential to form a true partitive (if the kind noun is definite).
Conversely, the NAC lacks this ability to form true partitives and has advanced further down the grammaticalisation path.
It is thus not surprising that it has lost (at least prototypically) the semantics which allows both the measure and the substance to be conceptually accessible as independent referents.
As a consequence, we should expect a tendency for the NAC constructions to host more strongly grammaticalised measure nouns.
For example, highly grammaticalised non-referential nouns like \textit{Gramm} `gram' and \textit{Meter} `metre' should occur proportionally more often in NAC constructions than in PGC constructions.
If such preferences can actually be shown in usage patterns, it would lend strong support for the hypothesised difference in the meanings of the prototypes.
In the corpus study, measure lemmas will therefore be annotated with appropriate semantic class labels to check whether semantic classes of measure nouns have different affinities to the two variants.
The actual classification used is based on the list in \citet[530]{Koptjevskaja2001}, but due to the low token frequencies in many of the potential classes, a very coarse classification was used in the end.
With typical examples, the classes are:
\textit{Physical} (typically non-referential precisely measurable units such as \textit{Liter} `litre', \textit{Meter} `metre', \textit{Gramm} `gram'),
\textit{Container} (\textit{Tasse} `cup'),
\textit{Amount} (\textit{Menge} `amount'), 
\textit{Portion} (natural portions like \textit{Happen} `bite' or \textit{Krümel} `crumb').
Notice that \textit{Container} is the class containing words like \textit{spoon} or \textit{cup}, which often develop into partially grammaticalised physical measure nouns.
The lemmas that did not fit into either of these classes were labelled \textit{Rest}.
